Taking Linear Logic Apart

### 2018-05-21 doc/abstract/main.tex
Finished first rewrite
Minor changes

\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{Linearity and TLLA 2018}
\usepackage[british]{babel}
\usepackage[all]{foreign}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\allowdisplaybreaks
\usepackage[inline]{enumitem}
\usepackage{breakurl} % Not needed if you use pdflatex only.
\usepackage{underscore} % Only needed if you use pdflatex.
\usepackage{parskip}
\usepackage{textcomp,textgreek,upgreek}
\usepackage{nameref,hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\hypersetup{hidelinks,final}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}[lemma]{Theorem}
\newtheorem{corollary}[lemma]{Corollary}
\newtheorem{definition}[lemma]{Definition}
\input{preamble-typing}
\usepackage[sectionbib,sort&compress]{natbib}

@@

In this paper, we introduce the Hypersequent Classical Processes (\hcp), which addresses this mismatch. The key insight is to register parallelism in the typing judgements using hypersequents~\citep{avron1991}, a technique from logic which generalises judgements from one sequent to many. This allows us to take apart the term constructs used in Classical Processes (\cp) to more closely match those of the standard \textpi-calculus. We proceed as follows. We start by introducing \cp (\cref{sec:cp}). Then, we introduce \hcp and prove it enjoys subject reduction and progress (\cref{sec:hcp}). We prove that every \cp process is an \hcp process, that every \hcp process corresponds to several entangled \cp processes, and that \hcp supports the same communication protocols as \cp, and no more (\cref{sec:cp2hcp}). Finally, we discuss related work (\cref{sec:related-work}).

@@

In this section, we discuss the relation between \cp and \hcp. We will prove three important theorems:
\item every \cp process is an \hcp process;
\item \hcp processes represent multiple entangled \cp processes; and
\item \hcp supports the same protocols as \cp.
To facilitate this, we define a translation from terms in \cp to terms in \hcp which breaks down the term constructs in \cp into their more atomic constructs in \hcp.
\begin{definition}\label{def:cp2hcp-terms}
    \begin{array}{lcllcl}
         \tm{\mtf{\cpLink{x}{y}}}
      &  := & \tm{\cpLink{x}{y}}
      &  \tm{\mtf{\cpCut{x}{P}{Q}}}
      &  := & \tm{\piNew{x}{(\piPar{\mtf{P}}{\mtf{Q}})}}
      \\ \tm{\mtf{\cpSend{x}{y}{P}{Q}}}
      &  := & \tm{\piBoundSend{x}{y}{(\piPar{\mtf{P}}{\mtf{Q}})}}
      &  \tm{\mtf{\cpRecv{x}{y}{P}}}
      &  := & \tm{\piRecv{x}{y}{\mtf{P}}}
      \\ \tm{\mtf{\cpHalt{x}}}
      &  := & \tm{\piBoundSend{x}{}{\piHalt}}
      &  \tm{\mtf{\cpWait{x}{P}}}
      &  := & \tm{\piRecv{x}{}{\mtf{P}}}
      \\ \tm{\mtf{\cpInl{x}{P}}}
      &  := & \tm{\cpInl{x}{\mtf{P}}}
      &  \tm{\mtf{\cpInr{x}{P}}}
      &  := & \tm{\cpInr{x}{\mtf{P}}}
      \\ \tm{\mtf{\cpCase{x}{P}{Q}}}
      &  := & \tm{\cpCase{x}{\mtf{P}}{\mtf{Q}}}
      &  \tm{\mtf{\cpAbsurd{x}}}
      &  := & \tm{\cpAbsurd{x}}
First, we prove that each \cp process can be translated by this trivial translation to an \hcp process, and that this translation respects structural congruence and reduction. Reductions from \cp can be trivially translated to reductions in \hcp. However, as we took apart each $\kappa$-rule into two separate rules, the reduction relation of \hcp is a slight refinement over that of \cp.
\begin{theorem}\label{thm:cp2hcp-typing}
  If $\seq[P]{\Gamma}$ in \cp, then $\seq[\mtf{P}]{\Gamma}$ in \hcp.
\end{theorem}\vspace*{-0.75\baselineskip}%
  By induction on the derivation of $\seq[P]{\Gamma}$. We show the interesting cases:
  \begin{itemize}
    Case \textsc{Cut}. We rewrite as follows:
    \begin{gather*}
      \begin{array}{lcl}
        \AXC{$\seq[P]{\Gamma, \tmty{x}{A}}$}
        \AXC{$\seq[Q]{\Delta, \tmty{x}{A^\bot}}$}
        \NOM{Cut}
        \BIC{$\seq[\cpCut{x}{P}{Q}]{\Gamma, \Delta}$}
        & \Rightarrow
        & \AXC{$\seq[\mtf{P}]{\Gamma, \tmty{x}{A}}$}
          \AXC{$\seq[\mtf{Q}]{\Delta, \tmty{x}{A^\bot}}$}
          \BIC{$\seq[\piPar{\mtf{P}}{\mtf{Q}}]{
          \Gamma, \tmty{x}{A} \hsep \Delta, \tmty{x}{A^\bot}}$}
          \UIC{$\seq[\piNew{x}{(\piPar{\mtf{P}}{\mtf{Q}})}]{
    \end{gather*}
    Case $(\tens)$. We rewrite as follows:
    \begin{gather*}
      \begin{array}{lcl}
        \AXC{$\seq[P]{\Gamma, \tmty{y}{A}}$}
        \AXC{$\seq[Q]{\Delta, \tmty{x}{B}}$}
        \BIC{$\seq[\cpSend{x}{y}{P}{Q}]{\Gamma, \Delta, \tmty{x}{A \tens B}}$}
        & \Rightarrow
        & \AXC{$\seq[\mtf{P}]{\Gamma, \tmty{y}{A}}$}
          \AXC{$\seq[\mtf{Q}]{\Delta, \tmty{x}{B}}$}
          \BIC{$\seq[\piPar{\mtf{P}}{\mtf{Q}}]{
          \Gamma, \tmty{y}{A} \hsep \Delta, \tmty{x}{B}}$}
          \UIC{$\seq[\piBoundSend{x}{y}{(\piPar{\mtf{P}}{\mtf{Q}})}]{
          \Gamma, \Delta, \tmty{x}{A \tens B}}$}
    \end{gather*}
    Case $(\one)$. We rewrite as follows:
    \begin{gather*}
      \begin{array}{lcl}
        \AXC{}\SYM{\one}
        \UIC{$\seq[\cpHalt{x}]{\tmty{x}{\one}}$}
        & \Rightarrow
        & \AXC{}
          \NOM{H-Halt}
          \UIC{$\seq[\piHalt]{\emptyhypercontext}$}
          \SYM{\one}
          \UIC{$\seq[\piBoundSend{x}{}{\piHalt}]{\tmty{x}{\one}}$}
    \end{gather*}
  \end{itemize}
  \vspace*{-\baselineskip}%
\end{proof}%
\begin{theorem}\label{thm:cp2hcp-equiv}
  $\tm{P}\equiv\tm{Q}$ in \cp iff $\tm{\mtf{P}}\equiv\tm{\mtf{Q}}$ in \hcp.
\end{theorem}\vspace*{-0.75\baselineskip}%
  By induction on the derivation of $\tm{P}\equiv\tm{Q}$.
\end{proof}%
\begin{theorem}\label{thm:cp2hcp-reduction}
  If $\reducesto{P}{Q}$ in \cp, then $\tm{\mtf{P}}\Longrightarrow^\star\tm{\mtf{Q}}$ in \hcp.
\end{theorem}\vspace*{-0.75\baselineskip}%
  By induction on the the derivation of $\reducesto{P}{Q}$.
\end{proof}%
\begin{theorem}\label{hcp2cp-reduction}
  If $\reducesto{\mtf{P}}{R}$ in \hcp, then there is a $\tm{Q}$ such that $\reducesto{P}{Q}$ in \cp and $\tm{R}\Longrightarrow^\star\tm{\mtf{Q}}$ in \hcp.
\end{theorem}\vspace*{-0.75\baselineskip}%
  By induction on the derivation of $\reducesto{\mtf{P}}{R}$. The cases for \hccpRedAxCut1 and the $\beta$-rules are trivial. For the $\kappa\nu$-rules, we rewrite by the appropriate $\kappa{\ppar}$ rule, and vice versa.

\hcp exhibits some behaviours which are impossible to directly translate back into \cp. For instance, in the process below, the choice sent on $\tm{x}$ will affect the choice between $\tm{P}$ and $\tm{P'}$, even though neither has access to the channel $\tm{x}$.
  \seq[\cpCase{x}{(\piPar{P}{Q})}{(\piPar{P'}{Q'})}]{
    \Gamma, \Delta, \tmty{x}{A \with B}}
Instead, we define a new relation to characterise the relation between proofs in \hcp and \cp.

@@

  Disentanglement is described by the smallest relation $\moveMixDown$ on proof derivations closed under the rules below.
  The relation $\moveMixDown^\star$ is the reflexive, transitive closure of $\moveMixDown$.

@@

@@

      \multicolumn{3}{l}{%

@@

      }

@@

    \AXC{}\noLine\UIC{$\rho_1$}

@@

    \AXC{}\noLine\UIC{$\rho_n$}

@@

Finally, we will prove that \hcp supports the same communication protocols as \cp. This is the same as saying that it inhabits the same session types, or that the associated logical systems derive the same theorems. We show this by proving that we can interpret the hyper-environments as formulas in the logic. This is a standard method for proving the soundness of a hypersequent calculus.

Environments can be interpreted as types by collapsing them as series of pars.
\begin{definition}\label{def:bigparr}
  \[
  \begin{array}{lcll}
    \ty{\bigparr(\emptycontext)}
    & = & \ty{\bot}
    \\
    \ty{\bigparr(\tmty{x_1}{A_1} , \dots , \tmty{x_n}{A_n})} 
     & = & \ty{A_1 \parr \dots \parr A_n}
     & \text{ if } n \geq 1
    \\
  \end{array}
  \]
\end{definition}\noindent
\begin{lemma}\label{lem:cp-bigparr}
  If $\seq{\Gamma}$ in \cp, then $\seq{\bigparr\Gamma}$ in \cp.
\end{lemma}\vspace*{-0.75\baselineskip}%
\begin{proof}
  By repeated application of $(\parr)$.
\end{proof}\noindent
Furthermore, hyper-environments can be interpreted as types by collapsing them as series of tensors where environments are handled using \cref{lem:hcp-disentangle}. The empty hyper-environment $\ty{\emptyhypercontext}$ (that is the unit of $\hsep$) is interpreted as the unit of tensor.
\begin{definition}\label{def:bigtens}
  \[
  \begin{array}{lcll}
    \ty{\bigtens(\emptyhypercontext)}
    & = & \ty{\one}
    \\
    \ty{\bigtens(\Gamma_1 \hsep \dots \hsep \Gamma_n)}
    & = & \ty{\bigparr\Gamma_1 \tens \dots \tens \bigparr\Gamma_n}
    & \text{ if } n \geq 1
  \end{array}
  \]
\end{definition}\noindent

>>>>

\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{Linearity and TLLA 2018}
\usepackage[british]{babel}
\usepackage[all]{foreign}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\allowdisplaybreaks
\usepackage[inline]{enumitem}
\usepackage{breakurl} % Not needed if you use pdflatex only.
\usepackage{underscore} % Only needed if you use pdflatex.
%\usepackage{parskip}
\usepackage{textcomp,textgreek,upgreek}
\usepackage{nameref,hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\hypersetup{hidelinks,final}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}[lemma]{Theorem}
\newtheorem{corollary}[lemma]{Corollary}
\newtheorem{definition}[lemma]{Definition}
\input{preamble-typing}
\usepackage[sectionbib,sort&compress]{natbib}

@@

In this paper, we introduce the Hypersequent Classical Processes (\hcp), which addresses this mismatch. The key insight is to register parallelism in the typing judgements using hypersequents~\citep{avron1991}, a technique from logic which generalises judgements from one sequent to many. This allows us to take apart the term constructs used in Classical Processes (\cp) to more closely match those of the standard \textpi-calculus. We proceed as follows. We start by introducing \cp (\cref{sec:cp}). Then, we introduce \hcp and prove it enjoys subject reduction and progress (\cref{sec:hcp}). We prove that every \cp process is an \hcp process, and that \hcp supports the same communication protocols as \cp, and no more (\cref{sec:cp2hcp}). Finally, we discuss related work (\cref{sec:related-work}).

@@

In this section, we discuss the relation between \cp and \hcp. We will prove two important theorems: every \cp process is an \hcp process; and \hcp supports the same protocols as \cp. To facilitate this, we define a translation from terms in \cp to terms in \hcp which breaks down the term constructs in \cp into their more atomic constructs in \hcp.
\begin{definition}\label{def:cp2hcp-terms}
  \[
    \begin{array}{lcllcl}
         \tm{\mtf{\cpLink{x}{y}}}
      &  := & \tm{\cpLink{x}{y}}
      &  \tm{\mtf{\cpCut{x}{P}{Q}}}
      &  := & \tm{\piNew{x}{(\piPar{\mtf{P}}{\mtf{Q}})}}
      \\ \tm{\mtf{\cpSend{x}{y}{P}{Q}}}
      &  := & \tm{\piBoundSend{x}{y}{(\piPar{\mtf{P}}{\mtf{Q}})}}
      &  \tm{\mtf{\cpRecv{x}{y}{P}}}
      &  := & \tm{\piRecv{x}{y}{\mtf{P}}}
      \\ \tm{\mtf{\cpHalt{x}}}
      &  := & \tm{\piBoundSend{x}{}{\piHalt}}
      &  \tm{\mtf{\cpWait{x}{P}}}
      &  := & \tm{\piRecv{x}{}{\mtf{P}}}
      \\ \tm{\mtf{\cpInl{x}{P}}}
      &  := & \tm{\cpInl{x}{\mtf{P}}}
      &  \tm{\mtf{\cpInr{x}{P}}}
      &  := & \tm{\cpInr{x}{\mtf{P}}}
      \\ \tm{\mtf{\cpCase{x}{P}{Q}}}
      &  := & \tm{\cpCase{x}{\mtf{P}}{\mtf{Q}}}
      &  \tm{\mtf{\cpAbsurd{x}}}
      &  := & \tm{\cpAbsurd{x}}
    \end{array}
  \]
\end{definition}\noindent
We will use this relation in the first proof, and its analogue for derivations in the second.

\subsection{Every \cp process is an \hcp process}
First, we prove that each \cp process can be translated by this trivial translation to an \hcp process, and that this translation respects structural congruence and reduction. Reductions from \cp can be trivially translated to reductions in \hcp. However, as we took apart each $\kappa$-rule into two separate rules, the reduction relation of \hcp is a slight refinement over that of \cp.
\begin{theorem}\label{thm:cp2hcp-typing}
  If $\seq[P]{\Gamma}$ in \cp, then $\seq[\mtf{P}]{\Gamma}$ in \hcp.
\end{theorem}\vspace*{-0.75\baselineskip}%
\begin{proof}
  By induction on the derivation of $\seq[P]{\Gamma}$. We show the interesting cases:
  \begin{itemize}
  \item
    Case \textsc{Cut}. We rewrite as follows:
    \begin{gather*}
      \begin{array}{lcl}
        \AXC{$\seq[P]{\Gamma, \tmty{x}{A}}$}
        \AXC{$\seq[Q]{\Delta, \tmty{x}{A^\bot}}$}
        \NOM{Cut}
        \BIC{$\seq[\cpCut{x}{P}{Q}]{\Gamma, \Delta}$}
        \DisplayProof
        & \Rightarrow
        & \AXC{$\seq[\mtf{P}]{\Gamma, \tmty{x}{A}}$}
          \AXC{$\seq[\mtf{Q}]{\Delta, \tmty{x}{A^\bot}}$}
          \NOM{H-Mix}
          \BIC{$\seq[\piPar{\mtf{P}}{\mtf{Q}}]{
          \Gamma, \tmty{x}{A} \hsep \Delta, \tmty{x}{A^\bot}}$}
          \NOM{H-Cut}
          \UIC{$\seq[\piNew{x}{(\piPar{\mtf{P}}{\mtf{Q}})}]{
          \Gamma, \Delta}$}
          \DisplayProof
      \end{array}
    \end{gather*}
  \item
    Case $(\tens)$. We rewrite as follows:
    \begin{gather*}
      \begin{array}{lcl}
        \AXC{$\seq[P]{\Gamma, \tmty{y}{A}}$}
        \AXC{$\seq[Q]{\Delta, \tmty{x}{B}}$}
        \SYM{\tens}
        \BIC{$\seq[\cpSend{x}{y}{P}{Q}]{\Gamma, \Delta, \tmty{x}{A \tens B}}$}
        \DisplayProof
        & \Rightarrow
        & \AXC{$\seq[\mtf{P}]{\Gamma, \tmty{y}{A}}$}
          \AXC{$\seq[\mtf{Q}]{\Delta, \tmty{x}{B}}$}
          \NOM{H-Mix}
          \BIC{$\seq[\piPar{\mtf{P}}{\mtf{Q}}]{
          \Gamma, \tmty{y}{A} \hsep \Delta, \tmty{x}{B}}$}
          \SYM{\tens}
          \UIC{$\seq[\piBoundSend{x}{y}{(\piPar{\mtf{P}}{\mtf{Q}})}]{
          \Gamma, \Delta, \tmty{x}{A \tens B}}$}
          \DisplayProof
      \end{array}
    \end{gather*}
  \item
    Case $(\one)$. We rewrite as follows:
    \begin{gather*}
      \begin{array}{lcl}
        \AXC{}\SYM{\one}
        \UIC{$\seq[\cpHalt{x}]{\tmty{x}{\one}}$}
        \DisplayProof
        & \Rightarrow
        & \AXC{}
          \NOM{H-Halt}
          \UIC{$\seq[\piHalt]{\emptyhypercontext}$}
          \SYM{\one}
          \UIC{$\seq[\piBoundSend{x}{}{\piHalt}]{\tmty{x}{\one}}$}
          \DisplayProof
      \end{array}
    \end{gather*}
  \end{itemize}
  \vspace*{-\baselineskip}%
\end{proof}%
\begin{theorem}\label{thm:cp2hcp-equiv}
  $\tm{P}\equiv\tm{Q}$ in \cp iff $\tm{\mtf{P}}\equiv\tm{\mtf{Q}}$ in \hcp.
\end{theorem}\vspace*{-0.75\baselineskip}%
\begin{proof}
  By induction on the derivation of $\tm{P}\equiv\tm{Q}$.
\end{proof}%
\begin{theorem}\label{thm:cp2hcp-reduction}
  If $\reducesto{P}{Q}$ in \cp, then $\tm{\mtf{P}}\Longrightarrow^\star\tm{\mtf{Q}}$ in \hcp.
\end{theorem}\vspace*{-0.75\baselineskip}%
\begin{proof}
  By induction on the the derivation of $\reducesto{P}{Q}$.
\end{proof}%
\begin{theorem}\label{hcp2cp-reduction}
  If $\reducesto{\mtf{P}}{R}$ in \hcp, then there is a $\tm{Q}$ such that $\reducesto{P}{Q}$ in \cp and $\tm{R}\Longrightarrow^\star\tm{\mtf{Q}}$ in \hcp.
\end{theorem}\vspace*{-0.75\baselineskip}%
\begin{proof}
  By induction on the derivation of $\reducesto{\mtf{P}}{R}$. The cases for \hccpRedAxCut1 and the $\beta$-rules are trivial. For the $\kappa\nu$-rules, we rewrite by the appropriate $\kappa{\ppar}$ rule, and vice versa.
\end{proof}\noindent

\subsection{\hcp supports the same communication protocols as \cp}
\hcp exhibits some behaviours which are impossible to directly translate back into \cp. For instance, in the process below, the choice sent on $\tm{x}$ will affect the choice between $\tm{P}$ and $\tm{P'}$, even though neither has access to the channel $\tm{x}$.
\[
  \seq[\cpCase{x}{(\piPar{P}{Q})}{(\piPar{P'}{Q'})}]{
    \Gamma, \Delta, \tmty{x}{A \with B}}
\]

Instead, we will prove that \hcp supports the same communication protocols as \cp. This is the same as saying that it inhabits the same session types, or that the associated logical systems derive the same theorems. We show this by proving that we can interpret the hyper-environments as formulas in the logic. This is a standard method for proving the soundness of a hypersequent calculus. We start off by defining a relation on derivations of \hcp.

@@

  Disentanglement is described by the smallest relation $\moveMixDown$ on proof derivations closed under the rules below, plus the associativity and commutativity of mixes. The relation $\moveMixDown^\star$ is the reflexive, transitive closure of $\moveMixDown$.

@@

    \setlength{\arraycolsep}{2pt}

@@

@@

@@

    \AXC{}\noLine\UIC{$\mtf{\rho_1}$}

@@

    \AXC{}\noLine\UIC{$\mtf{\rho_n}$}

@@

Environments can be interpreted as types by collapsing them as series of pars.
\begin{definition}\label{def:bigparr}
  \begin{array}{lcll}
    \ty{\bigparr(\emptycontext)}
    & = & \ty{\bot}
    \ty{\bigparr(\tmty{x_1}{A_1} , \dots , \tmty{x_n}{A_n})} 
     & = & \ty{A_1 \parr \dots \parr A_n}
     & \text{ if } n \geq 1
\begin{lemma}\label{lem:cp-bigparr}
  If $\seq{\Gamma}$ in \cp, then $\seq{\bigparr\Gamma}$ in \cp.
  By repeated application of $(\parr)$.
Furthermore, hyper-environments can be interpreted as types by collapsing them as series of tensors where environments are handled using \cref{lem:hcp-disentangle}. The empty hyper-environment $\ty{\emptyhypercontext}$ (that is the unit of $\hsep$) is interpreted as the unit of tensor.
\begin{definition}\label{def:bigtens}
  \begin{array}{lcll}
    \ty{\bigtens(\emptyhypercontext)}
    & = & \ty{\one}
    \ty{\bigtens(\Gamma_1 \hsep \dots \hsep \Gamma_n)}
    & = & \ty{\bigparr\Gamma_1 \tens \dots \tens \bigparr\Gamma_n}
    & \text{ if } n \geq 1
